---
title: "CENPC Screen - HCT116 - sgRNA Epigenetics Plus - downstream_Analysis"
author: "Krishnendu Guin/ Adib Keikhosravi / Gianluca Pegoraro"
date: today
date-format: long
format: 
  html:
      self-contained: true
      code-fold: true
      code-tools: true
      code-link: true
editor: source
---

## README

This script is used to further process the cellHTS2 results for downstream analysis. All the variables output from cellHTS2 analysis are collated in a single table for further analysis and plotting.

## Setup

Load packages.

```{r}
#| label: load-libraries
#| message: false
#| warning: false

library(tidyverse)
library(ggthemes)
library(fs)
```

```{r}
#| label: set-theme
#| include: false
#| warning: false

theme_set(theme_bw())
theme_update(
  axis.text.x = element_text(
    angle = -90,
    hjust = 0,
    vjust = 0.5
  ),
  axis.text.y = element_text(hjust = 0)
)
```

```{r}
here::i_am("Epiplus_sgrna_screen_HCT116/03_hts2_results_plotting.qmd")
```

## Read `cellHTS2` results

Read the `cellHTS2` results from the `hts2_output` directory.

```{r}
library(here)
read_hts2_results <- function(path, glob) {
  dir_ls(
    path = path,
    recurse = T,
    glob = glob
  ) |>
    map_df(read_tsv,
      .id = "file_name"
    )
}

glob_path <- "*Results_table.txt"
hts2_output_path <- here("Epiplus_sgrna_screen_HCT116", "hts2_output")

all_tbl <- read_hts2_results(hts2_output_path, glob_path) |>
  filter(!(is.na(score))) |>
  mutate(prop = str_match(file_name, "hts2_output/(.*?)/")[, 2]) |>
  select(plate,
    well,
    control = wellAnno,
    gene_symbol = GeneSymbol,
    prop,
    score,
    raw_repeat_1 = raw_r1_ch1,
    raw_repeat_2 = raw_r2_ch1,
    ratio_repeat_1 = `raw/PlateMedian_r1_ch1`,
    ratio_repeat_2 = `raw/PlateMedian_r2_ch1`,
    norm_repeat1 = normalized_r1_ch1,
    norm_repeat2 = normalized_r2_ch1
  ) |>
  pivot_wider(
    names_from = prop,
    values_from = c(
      score,
      raw_repeat_1,
      raw_repeat_2,
      ratio_repeat_1,
      ratio_repeat_2,
      norm_repeat1,
      norm_repeat2
    )
  ) |>
  rowwise() |>
  mutate(
    sd_cell_n = sd(
      c(
        norm_repeat1_cell_n,
        norm_repeat2_cell_n
      ),
      na.rm = T
    ),
    sd_k_area_mean = sd(
      c(
        norm_repeat1_area_mean,
        norm_repeat2_area_mean
      ),
      na.rm = T
    ),
    sd_k_corr_perc_mean = sd(
      c(
        norm_repeat1_k_corr_perc_mean,
        norm_repeat2_k_corr_perc_mean
      ),
      na.rm = T
    ),
    sd_radial_dist_mean_mean = sd(
      c(
        norm_repeat1_radial_dist_mean_mean,
        norm_repeat2_radial_dist_mean_mean
      ),
      na.rm = T
    ),
    sd_spots_number_mean = sd(
      c(
        norm_repeat1_spots_number_mean,
        norm_repeat2_spots_number_mean
      ),
      na.rm = T
    )
  ) |>
  ungroup() |>
  mutate(cell_line = "HCT116") |>
  relocate(cell_line) |>
  arrange(desc(score_k_corr_perc_mean))

glimpse(all_tbl)
```

Write the results to a csv file.

```{r}
write_csv(all_tbl, here("Epiplus_sgrna_screen_HCT116", "knitr_output", "hct116_all_genes.csv"))
```

```{r}
#|label: plot-cell-n-repeats

cell_n_repeats <- all_tbl |>
  ggplot(aes(
    x = norm_repeat1_cell_n,
    y = norm_repeat2_cell_n
  ))

cell_n_repeats +
  geom_point(
    shape = 21,
    aes(color = control)
  ) +
  geom_smooth(
    method = "lm",
    color = "grey50",
    se = F
  ) +
  coord_equal() +
  scale_color_few(name = "Treatment") +
  xlab("z-score Repeat 1") +
  ylab("z-score Repeat 2") +
  ggtitle("Cell Number")
```

```{r}
#|label: plot-area-repeats

area_repeats <- all_tbl |>
  ggplot(aes(
    x = norm_repeat1_area,
    y = norm_repeat2_area
  ))

cell_n_repeats +
  geom_point(
    shape = 21,
    aes(color = control)
  ) +
  geom_smooth(
    method = "lm",
    color = "grey50",
    se = F
  ) +
  coord_equal() +
  scale_color_few(name = "Treatment") +
  xlab("z-score Repeat 1") +
  ylab("z-score Repeat 2") +
  ggtitle("Nucleus Area")
```

```{r}
#|label: spot-n-repeats

spot_n_repeats <- all_tbl |>
  ggplot(aes(
    x = norm_repeat1_spots_number_mean,
    y = norm_repeat2_spots_number_mean
  ))

spot_n_repeats +
  geom_point(
    shape = 21,
    aes(color = control)
  ) +
  geom_smooth(
    method = "lm",
    color = "grey50",
    se = F
  ) +
  coord_equal() +
  scale_color_few(name = "Treatment") +
  xlab("z-score Repeat 1") +
  ylab("z-score Repeat 2") +
  ggtitle("CENPC Spots Number")
```

```{r}
#|label: k-repeats

k_repeats <- all_tbl |>
  ggplot(aes(
    x = norm_repeat1_k_corr_perc_mean,
    y = norm_repeat2_k_corr_perc_mean
  ))

k_repeats +
  geom_point(
    shape = 21,
    aes(color = control)
  ) +
  geom_smooth(
    method = "lm",
    color = "grey50",
    se = F
  ) +
  coord_equal() +
  scale_color_few(name = "Treatment") +
  xlab("z-score Repeat 1") +
  ylab("z-score Repeat 2") +
  ggtitle("% of Ripley's K curve above Poisson")
```

```{r cell_n-vs-spot_n, echo = F}
#| label: cell_n-vs-spot_n

cell_n_spot_n <- all_tbl |>
  ggplot(aes(
    x = score_cell_n,
    y = score_spots_number_mean
  ))

cell_n_spot_n +
  geom_point(
    shape = 21,
    aes(color = control)
  ) +
  scale_color_few(name = "Treatment") +
  xlab("Mean z-score Cell Number") +
  ylab("Mean z-score CENPC Spots") +
  ggtitle("Cell Number vs. CENPC Spots Number")
```

```{r}
#| label: cell_n-vs-ripley-k

cell_n_ripley_k <- all_tbl |>
  ggplot(aes(
    x = score_cell_n,
    y = score_k_corr_perc_mean
  ))

cell_n_ripley_k +
  geom_point(
    shape = 21,
    aes(color = control)
  ) +
  scale_color_few(name = "Treatment") +
  xlab("Mean z-score Cell Number") +
  ylab("Mean z-score Ripley's K") +
  ggtitle("Cell Number vs. Ripley's K")
```

```{r}
#| label: spot-n-vs-ripley-k

spot_n_ripley_k <- all_tbl |>
  ggplot(aes(
    x = score_spots_number_mean,
    y = score_k_corr_perc_mean
  ))

spot_n_ripley_k +
  geom_point(
    shape = 21,
    aes(color = control)
  ) +
  scale_color_few(name = "Treatment") +
  xlab("Mean z-score CENPC Spots") +
  ylab("Mean z-score Ripley's K") +
  ggtitle("CENPC Spots Number vs. Ripley's K")
```

```{r}
#| label: area-vs-ripley-k

spot_n_ripley_k <- all_tbl |>
  ggplot(aes(
    x = score_area_mean,
    y = score_k_corr_perc_mean
  ))

spot_n_ripley_k +
  geom_point(
    shape = 21,
    aes(color = control)
  ) +
  scale_color_few(name = "Treatment") +
  xlab("Mean z-score Nucleus Area") +
  ylab("Mean z-score Ripley's K") +
  ggtitle("Nucleus Area vs. Ripley's K")
```

```{r}
samples <- all_tbl |>
  filter(control == "sample")

glimpse(samples)
```

Filter putative hits based on Z-score for Ripley's K above or below +/-3, no cytotoxicity, and limited variance between biological replicates.

```{r}
#| label: putative-hits-k

putative_hits_k <- samples |>
  filter(
    score_k_corr_perc_mean > 3 | score_k_corr_perc_mean < -3,
    score_cell_n > -2,
    abs(score_k_corr_perc_mean) > sd_k_corr_perc_mean,
  ) |>
  select(
    plate:gene_symbol,
    score_k_corr_perc_mean,
    sd_k_corr_perc_mean,
    score_spots_number_mean,
    sd_spots_number_mean,
    score_cell_n,
    sd_cell_n,
    score_radial_dist_mean_mean,
    sd_radial_dist_mean_mean
  ) |>
  arrange(desc(score_k_corr_perc_mean))

rmarkdown::paged_table(putative_hits_k)
```

```{r}
#| label: putative-hits-n

putative_hits_n <- samples |>
  filter(
    score_spots_number_mean > 3 | score_spots_number_mean < -3,
    score_cell_n > -2,
    abs(score_spots_number_mean) > sd_spots_number_mean,
  ) |>
  select(
    plate:gene_symbol,
    score_k_corr_perc_mean,
    sd_k_corr_perc_mean,
    score_spots_number_mean,
    sd_spots_number_mean,
    score_cell_n,
    sd_cell_n,
    score_radial_dist_mean_mean,
    sd_radial_dist_mean_mean
  ) |>
  arrange(desc(score_spots_number_mean))

rmarkdown::paged_table(putative_hits_n)
```

```{r}
#| label: putative-hits-r

putative_hits_r <- samples |>
  filter(
    score_radial_dist_mean_mean > 3 | score_radial_dist_mean_mean < -3,
    score_cell_n > -2,
    abs(score_radial_dist_mean_mean) > sd_radial_dist_mean_mean,
  ) |>
  select(
    plate:gene_symbol,
    score_k_corr_perc_mean,
    sd_k_corr_perc_mean,
    score_spots_number_mean,
    sd_spots_number_mean,
    score_cell_n,
    sd_cell_n,
    score_radial_dist_mean_mean,
    sd_radial_dist_mean_mean
  ) |>
  arrange(desc(score_radial_dist_mean_mean))

rmarkdown::paged_table(putative_hits_r)
```

Document the information about the analysis session

```{r}
#| label: session-info
#| results: markup

sessionInfo()
```
